\huge Inferring microbial co-occurrence networks from 16S data: A systematic evaluation

### 2022-12-01 final_submission/manuscript.tex
Add: Draft for final submission
Change: Bootstrap -> Permutation in the manuscript

  \usepackage{hyperref} % For hyperlinks in the PDF
  \usepackage{csquotes}
  \usepackage[style=nature,url=false,backend=biber,sorting=none]{biblatex}
  \bibliography{references/references.bib}

@@

  The pipeline includes 4 methods for pairwise correlation metrics, and 6 methods for direct association metrics (refer to Table~\ref{tab:micone_tools}).
  Pairwise correlation methods involve the calculation of the correlation coefficient between each pair of nodes (taxonomic entity like Genera) leading to the inclusion of spurious indirect connections.
  On the other hand, direct association methods use conditional independence to avoid the detection of correlated but indirectly connected taxonomic entities~\cite{Kurtz2015,Menon2018}.
  A null model is created by re-sampling and bootstrapping the counts matrix and recalculating the correlations (see next section for details on network analysis and statistics).
  These bootstrapped association matrices are used to calculate the significance of the inferred correlations by calculating the p-values against this null model~\cite{Watts2018}.
  Brown's p-value merging method~\cite{brown_400_1975} is used for combining p-values from the pairwise correlations methods to obtain a consensus p-value, which can be used to filter for significance.
  The bootstrapping and p-value calculations are only performed on the correlations-based methods.
  In the final module of this step, the consensus algorithms are used to create the final consensus network using associations from all the network inference methods (except Pearson and Spearman, by default).
  The outputs of this step are co-occurrence association networks encoded in the JSON format (refer to Supplementary section) and which can also be exportable to a variety of network formats.
  The list of tools used in this step, along with their modules and references are provided in Table~\ref{tab:micone_tools}.

@@

  \subsubsection*{Bootstrapping and p-value calculation}
  \vspace{-5mm}
  For all correlation-based methods $k \le w$, $1000$ permutations of the original \ac{otu} counts data were generated~\cite{Watts2018}.
  The correlations in the permuted \ac{otu} tables are recalculated using the different correlation-based algorithms.
  Finally, the p-value is determined based on how often a more extreme association is observed for randomly permuted data.
  Note that, all the direct association-based methods used in the study have their own regularization methods built in and hence do not need to undergo bootstrapping.

@@

  \begin{table}[H]
    \centering
    \caption{
      \textbf{Tools used in the \ac{micone} pipeline}.
      The tools highlighted in gray are the defaults for the pipeline that are recommended based on the benchmarks with the mock and synthetic datasets.
      The consensus algorithm in the Network Inference (\ac{ni}) step incorporates all the modules (bootstrapping, direct association, and correlation-based) to generate the consensus network.
    }
    \label{tab:micone_tools}
  \end{table}

>>>>

  \usepackage{hyperref} % For hyperlinks in the PDF
  \usepackage{csquotes}
  \usepackage[style=nature,url=false,backend=biber,sorting=none]{biblatex}
  \bibliography{references.bib}

@@

  The pipeline includes 4 methods for pairwise correlation metrics, and 6 methods for direct association metrics (refer to Table~\ref{tab:micone_tools}).
  Pairwise correlation methods involve the calculation of the correlation coefficient between each pair of nodes (taxonomic entity like Genera) leading to the inclusion of spurious indirect connections.
  On the other hand, direct association methods use conditional independence to avoid the detection of correlated but indirectly connected taxonomic entities~\cite{Kurtz2015,Menon2018}.
  A null model is created by re-sampling and permuting the counts matrix and recalculating the correlations (see next section for details on network analysis and statistics).
  These permuted association matrices are used to calculate the significance of the inferred correlations by calculating the p-values against this null model~\cite{Watts2018}.
  Brown's p-value merging method~\cite{brown_400_1975} is used for combining p-values from the pairwise correlations methods to obtain a consensus p-value, which can be used to filter for significance.
  The permutations and p-value calculations are only performed on the correlations-based methods.
  In the final module of this step, the consensus algorithms are used to create the final consensus network using associations from all the network inference methods (except Pearson and Spearman, by default).
  The outputs of this step are co-occurrence association networks encoded in the JSON format (refer to Supplementary section) and which can also be exportable to a variety of network formats.
  The list of tools used in this step, along with their modules and references are provided in Table~\ref{tab:micone_tools}.

@@

  \subsubsection*{Permutations and p-value calculation}
  \vspace{-5mm}
  For all correlation-based methods $k \le w$, $1000$ permutations of the original \ac{otu} counts data were generated~\cite{Watts2018}.
  The correlations in the permuted \ac{otu} tables are recalculated using the different correlation-based algorithms.
  Finally, the p-value is determined based on how often a more extreme association is observed for randomly permuted data.
  Note that, all the direct association-based methods used in the study have their own regularization methods built in and hence do not need to undergo this procedure.

@@

  \begin{table}[H]
    \centering
    \caption{
      \textbf{Tools used in the \ac{micone} pipeline}.
      The tools highlighted in gray are the defaults for the pipeline that are recommended based on the benchmarks with the mock and synthetic datasets.
      The consensus algorithm in the Network Inference (\ac{ni}) step incorporates all the modules (permutations, direct association, and correlation-based) to generate the consensus network.
    }
    \label{tab:micone_tools}
  \end{table}
